---
title: "ENVS 193DS HW2"
description: "A homework assignment from my Statistics for Environmental Science Class that covers multiple linear regression visualization, affective data visualization, and statistical critique."
author: 
  - name: Mateo Mori
    url: https://mateomoribaez-design.github.io/
date: 03-12-2026
categories: [ENVS 193DS, RStudio] 
url: https://mateomoribaez-design.github.io/posts/Homework-2/
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
---

---
title: "Homework 2 Quarto"
author: "Mateo Mori"
format: html
editor: visual
toc: true
---

```{r}
#| message: false
library(tidyverse) # load in packages
library(janitor)
```

```{r}
#| message: false
lobsters <- read_csv("lobster-catch-data.csv") # read in lobster data
my_data <- read_csv("my_data.csv")
```

## Part 2. Problems

### Problem 1

a.  I collected discrete data to represent the presence of the bunchgrass species across 13 different sites within the oak woodlands. I collected this type of data because it is not categorical and it only takes integer values.

b.  Standard deviation would be best because it reflects the actual spread of bunchgrass counts in the data. My final answer is 1.9 bunchgrass individuals (count) rounded to 1 decimal point.

```{r}
#| message: false
bunchgrass_count <- c (4, 6, 4, 5, 5, 4, 2, 3, 6, 4, 2, 0, 1) # saving list of numbers as an object called bunchgrass_count
sd(bunchgrass_count) # caluculated standard deviation
```

c\. Standard error is the better description of uncertainty because it shows the precision behind the bunchgrass count mean. The standard error in this scenatio would represent the average amount I would expect the sample mean to vary from the true population mean if we were to repeat the study multiple times with the same sample size (13). My final answer is 0.5 bunchgrass individuals rounded to 1 decimal point, and this means that our sample mean is relatively precise, and that the true population mean likely falls within a narrow range around our calculated average.

```{r}
#| message: false
sd(bunchgrass_count)/sqrt(length(bunchgrass_count)) # caluculated standard error
```

### Problem 2

#### 2a.

```{r}
#| message: false

lobsters_clean <- lobsters |> # introduced lobsters dataframe
  clean_names() |> # cleaned lobsters dataframe to only show lower case letters and underscores
  mutate(protection_type = case_when( #changed MPA and non-MPA labels to be Marine Protected Area and not protected, respectively.
    protection_type == "MPA" ~ "Marine Protected Area",
    protection_type == "non-MPA" ~ "not protected"
  )) |>
  select(protection_type, weight_lbs) # selected only these two columns to be shown in the data frame.

lobsters_clean |> # introduced lobster data
  slice_sample(n = 10) # displayed 10 random rows from the two columns of the data frame

```

#### 2b.

```{r}
#| message: false
lobster_summary <- lobsters_clean |> # created new object from lobsters_clean
  group_by(protection_type) |> # comparing protection types
  summarize( 
    mean = mean(weight_lbs),
    n = length(weight_lbs),
    sd = sd(weight_lbs)
    ) # calculated mean, standard deviation and recorded number of observation
lobster_summary

```

#### 2c.

```{r}
#| message: false
ggplot(lobsters_clean, # introduce lobsters_clean dataframe
       aes(x = protection_type, # x and y axis using their own column in the dataframe
           y = weight_lbs,
           color = protection_type, # different colors per protection type
           shape = protection_type # different shape per protection type
           )) + 
  geom_jitter( # show data in jitterplot format
    height = 0,
    width = 0.1,
    alpha = 0.6,
    size = 2 # change the width of the horizontal jitter spread, the transparency of the points, and the size of each of the points.
  ) +
  geom_pointrange(data = lobster_summary,
                  aes(x = protection_type, # x is protection type
                      y = mean, # y is mean
                      ymin = mean - sd, # lower bound
                      ymax = mean + sd), # upper bound
                  color = "black", # add black outline and decrease size a little
                  size = 0.5,
                  inherit.aes = FALSE) +
  scale_color_manual(values = c("steelblue", "forestgreen")) + # manually changes colors of the two different sample groups
  theme_minimal() + # changed theme to not be the default
  theme(legend.position = "none") + # removes legend
  labs( # creates axis labels
    x = "Protection Type",
    y = "Mean trapped lobster weight (lbs)"
  )
```

#### 2d.

Null Hypothesis: The mean trapped weight of California spiny lobster is the same in MPA and non-MPA sites

Alternative Hypothesis: The mean trapped weight of California spiny lobster differs between MPA and non-MPA sites

#### 2e.

```{r}
#| message: false
ggplot(data = lobsters_clean, # indroduced dataframe to be used in the plot
       aes(sample = weight_lbs)) + # made y - axis
  geom_qq_line(color = "blue") + # 
  geom_qq() + # create qq plot with the data introduced above
  facet_wrap(~ protection_type) # shows protection categories separately
```

#### 2f.

The variable of interest is normally distributed because the shape of the distribution falls on the reference line in the QQ plot. Not only does the general shape align with the reference line, but the few outliers in the individual lobster weights can be seen as not significant enough to indicate skewness, accepting the data as normally distributed. Both "Marine Protected Areas" and "not protected" groups show these patterns.

#### 2g.

Using the F-test result ($p = 0.29$), there is no significant difference in the variance of lobster weights between the Marine Protected Area and non-protected sites, so equal variance can be assumed between the two data sets. This means that the spread of data points around the mean are roughly the same across the two groups, backing up the claim of there being no significant difference.

```{r}
#| message: false
var.test(weight_lbs ~ protection_type, # compared the variance of the two normally distributed samples, seeing if they are equal or not.
         data = lobsters_clean) # introduced the dataframe used for the variance test
```

#### 2h.

```{r}
#| message: false
t.test(
  weight_lbs ~ protection_type, # formula: response variable ~ grouping variable
  var.equal = TRUE, # argument for equal/unequal variances (variances should be equal)
  data = lobsters_clean # data: lobsters data frame
)
```

#### 2i.

-   Using a t-test is appropriate for this situation because we are comparing mean lobster weights between two independent groups (MPA and non-protected) using a continuous response variable.

-   Normal distribution was evaluated through the data points following the reference in the QQ plot, and homogeneity of variance was assessed by using an F-test to compare two variances.

#### 2j.

We found a significant difference in mean lobster weight between MPA (n = 35) and non-protected areas (n = 30) (Students t-test, t(63) = -10.08, p \< 0.001, a = 0.05).
